Covariate analysis of MAYO reprocessed counts (with CQN normalisation and covariates adjusted)

### 2017-03-23 code/Rmd/MAYO_REPROCESSED.Rmd
Modified Mayo reprocessing to include modified algorithms for SVA and samples with picard reruns

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering, cache=TRUE}
primaryVariable <- "BrainRegion.Diagnosis"
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS", "PCT_PF_READS_ALIGNED",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp, fig.height=6, fig.width=18}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

@@

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
  group_by(Region, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05)

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
  group_by(Region, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2)

kable(full_join(tmp %>%
                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
                tmp1 %>%
                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))

@@

# Source modified functions from sva package
source('../R/num.sv.R')
source('../R/irwsva.build.R')
source('../R/f.pvalue.R')

# Calculate the number of surrogate variables
NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                     MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)

>>>>

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering, cache=TRUE}
primaryVariable <- "BrainRegion.Diagnosis"
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
  dplyr::mutate(Region = factor(Region),
                Source = factor(Source))
plotdata$SampleID[!(plotdata$SampleID %in% c("11311_CER", "1923_CER", "1923_TCX", "11396_TCX", "11294_TCX", "11408_TCX"))] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp, fig.height=6, fig.width=18, cache=FALSE}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

@@

tmp = DE %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
  group_by(Region, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

@@

# Source modified functions from sva package
# source('../R/num.sv.R')
source('../R/irwsva.build.R')
source('../R/f.pvalue.R')

# Calculate the number of surrogate variables
# NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
#                      MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)

# Get residuals from differential expression model
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)

# Compute actual variance of all principal components
tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
actual.var = tmp$d^2/sum(tmp$d^2)

# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])

# Plot variance components
var.comp = data.frame(component = 1:length(actual.var), 
                      residual = actual.var, 
                      permuted.residual = permuted.var) %>%
  tidyr::gather(data, value, -component)
p = ggplot(var.comp %>% filter(component <= 10), aes(x = component, y = value, color = data)) + geom_point()
p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
p
